Genetic Dilution erodes productivity: Exploring farmers' low adoption levels of improved maize in Ethiopia

### 2023-03-24 paper.tex
make changes

    \clearpage\maketitle
	\thispagestyle{empty}
	\vspace*{-2em}
    \begin{center}\begin{abstract}
    \singlespacing
			\noindent  
			Despite large investments in  developing new  maize varieties, evidence shows that few farmers consistently adopt this new technology over time and that average effects mask considerable hetergeneity in impacts. Misclassification of seed is a crucial problem in these contexts, as non-classical measurement error can bias the impacts. We take a step further and study misclassification in the context of heterogeneous impacts to adoption. We use a three-year panel to estimate heterogeneous returns to adoption from self-reports of hybrid seed use and find peculiar results driven by bias due to the misclassification of seed. We exploit unique information on maize DNA-fingerprinting collected over the same areas in 2018. We find that misclassification in self-reported hybrid seed adoption mask heterogeneity in the actual quality of the genetic material of the seeds farmers categorized as improved varieties in the self-reported data. Positive returns to adoption are found for those farmers that use higher-purity germplasm, drought-tolerant maize, and newly released varieties. When we account for misclassification in our heterogeneity analysis, we find that the puzzle is resolved. Our findings point to the wide dispersion of older and genetically diluted varieties, for which poorer farmers may be paying a premium. The implications of our findings speak to the need for policies to better target context and geography, expand accessibility of improved seeds, and make higher yields varieties more inclusive.

@@

Misclassification of improved seed and its implications for measuring productivity impacts has been a relatively new branch of study, with the advent of DNA fingerprinting data. For instance, \cite{wossen2019estimating} use DNA fingerprinting data to show that there is a 22\% bias in the productivity impact of improved cassava seed in Nigeria. In the context of hybrid seed adoption, however, there is evidence to suggest that productivity impacts of seed affect households heterogeneously \citep{Suri2011-oi}, and that average impacts to adoption may not be as informative. The observed effect of adoption may be low because households that would have a comparative advantage in adoption of improved seed are constrained by a lack of access to improved seed or infrastructure. Additionally, the non-classical measurement error introduced by misclassification can bias heterogeneity in ways that may attenuate or even reverse productivity impacts. This paper provides a way to account for misclassification of seed when additional, error-free data exists, such as DNA fingerprinting data. We estimate the comparative advantage of households directly using a Group Random Coefficients (GRC) Model and correct its heterogeneity estimates using misclassification rates from DNA fingerprinting data of the same plots \citep{Suri2011-oi, Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc}.

@@

Our results contribute to the literature on the drivers of low adoption rates of improved seed varieties. The consensus in this literature is that the success of improved varieties depends on a range of factors, and that the widespread adoption of  improved seeds is the result of complementary conditions and inputs that allow farmers to extract the full potential of these varieties. Specifically, extension services and on-farm field trials, seed variety characteristics and rainfall have been found to be crucial to the adoption of improved maize in Tanzania \citep{Kaliba2000-jh}. In the specific case of Ethiopia, labor, fertilizer use, farmers’ experience with extension packages, rainfall suitability and prices have been shown to increase the opportunity cost of not adopting improved wheat varieties \citep{Wale2006-bv}. Market access and the accessibility of extension services are also critical in the adoption of improved chickpea varieties in Ethiopia \citep{Verkaart2019-ol}. Other important determinants of heterogeneous effects include risk aversion \citep{Holden2016-vy}, farm size \citep{Ghimire2015-bd}, and credit constraints \citep{Simtowe2008-jn,Balana2020-hx}. We are the first to consider how heterogeneous impacts to seed adoption are affected by misclassification of seed.

@@

These findings may also hold for our case, especially taking into consideration the degree of genetic dilution and dispersion of recycled or older-generation improved varieties, which are not as yield responsive to fertilizer application. In addition to this, prior evidence also indicates that farmers sub-optimally allocate fertilizer in response to their mislead beliefs about the crop variety (\citealt{euler2022because}). Such wrong beliefs or misclassification has also been shown to be driven by misperception instead of misreporting (\citealt{wossen2022misperceiving}), which is also the case for the farmers in our sample. These findings suggest that the seemingly small to negative returns to adoption that we find through the GRC model are driven by both a lower response to fertilizers of the older-generation improved varieties, but also to farmers’ sub-optimal application of fertilizers (and possibly other inputs) due to misperception.

The suitability of the improved varieties to weather conditions and the weather conditions themselves are two additional factors that may further explain the seemingly negative comparative advantages to adopting the misperceived improved varieties. Even moderate droughts have been shown to decrease maize yields of non-drought tolerant varieties by 20\% (\citealt{paul2021heterogeneous}). Additionally, although our results confirm the positive returns to DTM adoption (consistent with \citealt{wossen2017measuring}), even the DTM varieties distributed in Ethiopia are only resistant to mild drought but would also fail like any other non-drought tolerant variety in case of severe droughts.

>>>>

    \clearpage\maketitle
	\thispagestyle{empty}
	\vspace*{-2em}
    \begin{center}\begin{abstract}
    \singlespacing
			\noindent  
			Despite large investments in developing new maize varieties, evidence shows that few farmers consistently adopt them. Average effects on yields mask considerable heterogeneity. Misclassification of seed is a crucial problem in these contexts, as non-classical measurement error can lead to biased effects. We study misclassification in the context of heterogeneous impacts to adoption. We use a three-year panel to estimate heterogeneous returns to adoption from self-reports of hybrid seed use and find results driven by bias due to the seed misclassification. We leverage unique information on maize DNA-fingerprinting collected over the same areas in 2018. Misclassification in self-reported hybrid seed adoption masks heterogeneity in the quality of the genetic material categorized as improved varieties. Accounting for seed misclassification, we find that heterogeneity in the actual quality of the genetic material masks positive returns to adoption, with the highest returns found for farmers using higher-purity germplasm, drought-tolerant maize, and newly released varieties. When we account for misclassification in our heterogeneity analysis, we find that farmers that have the largest potential to benefit from adoption are constrained in their ability to do so. The wide dispersion of older and genetically diluted varieties, for which poorer farmers may be paying a premium points to the need for policies that expand the accessibility of improved seed and verify its quality.

@@

% Given the ubiquity of improved seed varieties over the course of decades of access in Ethiopia,

Misclassification of improved seed and its implications for measuring productivity impacts has been a relatively new branch of study, with the advent of DNA fingerprinting data. For instance, \cite{wossen2019estimating} use such data to show that there is a 22\% bias in the productivity impact of improved cassava seed in Nigeria. Other literature suggests that hybrid seed adoption is heterogeneous across households \citep{Suri2011-oi}, and that average impacts to adoption may not be as informative. The observed effect of adoption may be low because households that would have a comparative advantage in adoption of improved seed are constrained by a lack of access to improved seed or necessary infrastructure. Additionally, non-classical measurement error introduced by misclassification can bias heterogeneity in ways that may attenuate or even reverse productivity impacts. This paper provides a way to account for misclassification of seed when additional, error-free data exists, such as DNA fingerprinting data. We estimate the comparative advantage of households directly using a Group Random Coefficients (GRC) Model and correct its heterogeneity estimates using misclassification rates from DNA fingerprinting data of the same plots \citep{Suri2011-oi, Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc}.

@@

The consensus in this literature is that the success of improved varieties depends on a range of factors, and that the widespread adoption of  improved seeds is the result of complementary conditions and inputs that allow farmers to extract the full potential of these varieties. In the specific case of Ethiopia, labor, fertilizer use, farmers’ experience with extension packages, rainfall suitability and prices have been shown to increase the opportunity cost of not adopting improved wheat varieties \citep{Wale2006-bv}. Market access and the accessibility of extension services are also critical in the adoption of improved chickpea varieties in Ethiopia \citep{Verkaart2019-ol}. Other important determinants of heterogeneous effects include risk aversion \citep{Holden2016-vy}, farm size \citep{Ghimire2015-bd}, and credit constraints \citep{Simtowe2008-jn,Balana2020-hx}. We are the first to consider how heterogeneous impacts to seed adoption are affected by misclassification of seed.\footnote{In other contexts, extension services and on-farm field trials, seed variety characteristics and rainfall have been found to be crucial to the adoption of improved maize in Tanzania \citep{Kaliba2000-jh}.}

@@

These findings may also hold for our case, especially taking into consideration the degree of genetic dilution and dispersion of recycled or older-generation improved varieties, which are not as yield responsive to fertilizer application. In addition to this, prior evidence also indicates that farmers sub-optimally allocate fertilizer in response to their mislead beliefs about the crop variety leading to lower yield realization and consequent dis-adoption  (\citealt{euler2022because}). Such wrong beliefs or misclassification has also been shown to be driven by misperception instead of misreporting (\citealt{wossen2022misperceiving}), which is also the case for the farmers in our sample. These findings suggest that the seemingly small to negative returns to adoption that we find through the GRC model are driven by a combination of lower responses to fertilizers of the older-generation improved varieties and farmers’ sub-optimal application of fertilizers (and possibly other inputs) due to misperception.

The suitability of the improved varieties to location-specific weather conditions as well as the variation in the weather conditions themselves are two additional factors that may further explain the seemingly negative comparative advantages to adopting the misperceived improved varieties. Even moderate droughts have been shown to decrease maize yields of non-drought tolerant varieties by 20\% (\citealt{paul2021heterogeneous}). Additionally, although our results confirm the positive returns to DTM adoption (consistent with \citealt{wossen2017measuring}), even the DTM varieties distributed in Ethiopia are only resistant to mild drought but would also fail like any other non-drought tolerant variety in case of severe droughts.
